Applications of Many-Core Technologies to On-line Event Reconstruction in High Energy Physics Experiments

### 2013-11-13 papers/2013/NSS/NSS.tex
some fixes

\documentclass[journal]{IEEEtran}
\usepackage{graphicx}
\usepackage{subfigure}

\usepackage{hyperref}
\usepackage{multirow}
\usepackage[all]{hypcap} % fix links to figs/tabs

@@

% disable the running header on all pages following the title page
\pagestyle{empty}
%
\begin{document}
% paper title
\title{Title of NSS Talk}
% Author list
\author{  A.~Gianelle, 
  S.~Amerio, 
  D.~Bastieri, 
  M.~Corvo, 
  W.~Ketchum,
  T.~Liu, 
  A.~Lonardo, 
  D.~Lucchesi,
  S.~Poprocki, 
  R.~Rivera, 
  L.~Tosoratto,
  P.~Vicini
  and 
  P.~Wittich
}
\thanks{Manuscript received November 15, 2013.
This work was supported by the National Science Foundation, the
Department of Energy Office of Science and INFN.}% <-this % stops a space
\thanks{A.~Gianelle, S.~Amerio, D.~Bastieri, M.~Corvo, and D.~Lucchesi are with INFN Padova.}% 
\thanks{W.~Ketchum is with Los Alamos National Laboratory.}%
\thanks{T.~Liu and R.~Rivera are with Fermi National Accelerator Laboratory.}%
\thanks{A.~Lonardo, L.~Tosoratto and P.~Vicini are with INFN Roma}% 
\thanks{S.~Poprocki and P.~Wittich are with Cornell University.}%
% make the title area
\maketitle
% after \maketitle, add the following command to remove the header on the title page
\thispagestyle{empty}
% abstract usually in passive voice
\begin{abstract}
  Interest in parallel architectures applied to real time selections is 
  growing in High Energy Physics (HEP) experiments. In this paper we describe
  performance measurements of Graphic Processing Units (GPUs) and Intel Many
  Integrated Core architecture (MIC) when applied to a typical HEP online task: 
  the selection of events based on the trajectories of charged particles.
  We use as benchmark a scaled-up version of the algorithm used at CDF 
  experiment at Tevatron for online track reconstruction - the SVT algorithm - 
  as a realistic  test-case for low-latency trigger systems using new computing
  architectures for LHC experiment. We examine the
  complexity/performance trade-off in porting existing serial
  algorithms to many-core devices. Measurements of both data processing 
  and data transfer latency are shown, considering different I/O strategies 
  to/from the parallel devices. \end{abstract}

\section{Introduction}
Real-time event reconstruction plays a fundamental role in High Energy
Physics (HEP) experiments at hadron colliders.  Reducing the rate of
data to be saved on tape from millions to hundreds of events per
second is critical. To increase the purity of the collected samples,
rate reduction has to be coupled with an initial selection of the most
interesting events.  In a typical hadron collider experiment, the
event rate has to be reduced from tens of MHz to a few kHz.  The
selection system (trigger) is usually organized in multiple levels,
each capable of performing a finer selection on more complex physics
objects describing the event. Trigger systems usually comprise a first
level based on custom hardware, followed by one or two levels usually
based on farms of general purpose processors.  At all levels, latency
is a concern: for a fixed processing time, the faster a decision is
rendered about accepting or rejecting an event improves the purity of
the collected data sample. The possibility of using commercial devices
at a low trigger level is very appealing: they are subject to
continuous performance improvements driven by the consumer market, are
less expensive than dedicated hardware, and are easier to support.
Among the commercial devices, many-core architectures such as Graphic
Processing Units (GPUs)~\cite{bib_gpu} and Intel Many Integrated Core
(MIC)~\cite{bib_intelMIC} are of particular interest for online
selections given their great computing power: the latest
\textsc{nvidia}~\cite{bib_nvidia} GPU architecture, Kepler, exceeds
Teraflop computing power. Moreover, high-level programming
architectures based on \textsc{c/c++} such as
\textsc{cuda}~\cite{bib_cuda} and \textsc{opencl}~\cite{bib_opencl}
make programming these devices more accessible to the general
physicist user.  The goal of this study is to investigate the
strengths and weaknesses of many-core devices when applied in a low
latency environment, with particular emphasis on the data transfer
latency to/from the device and the algorithm latency for processing on
the device in a manner similar to a typical HEP trigger application,
and to understand the cost/complexity ratio of porting legacy serial
code to many-core devices.

@@

\begin{table}[!t]
  \centering
  \begin{tabular}{|l|c|c|c|c|c|}
    \hline
    Model & Tesla M2050 & Tesla K20m & GeForce GTX  Titan & MIC 5110P & Radeon HD 7970 \\
    \hline
    \hline
    Performance (SP, GFlops) & 1030 & 3520 & 4500 & 2022 & 3790\\
    Memory bandwidth  (GB/s) & 148 & 208 & 288  & 320 & 264\\   
    Memory size (GB) & 3 & 5 & 6 & 8 & 3\\
    Number of cores & 448 & 2496 & 2688 & 240 & 2048\\
    Clock speed (GHz) & 1.15 & 0.706 & 0.837 & 1.053 & 1.375\\
    \hline
  \end{tabular}
  \caption{Capabilities of the many-core devices used in this study,
    according to the manufacturer's specifications. The first three
    are \textsc{nvidia} GPUs, the MIC 5110P is an Intel Xeon Phi, and the final one is an AMD GPU.}
  \label{tab_hwspecs}
\end{table}

To measure the data transfer latency we use a computing cluster composed of 
12 identical nodes.  Each node contains a Intel Xeon E6520 2.4 GHz 
CPU and two Tesla M2075 GPU cards. The nodes are connected by InfiniBand 
communication links using Connect-X2 Mellanox or APEnet+ adapters. 
APEnet+ is an FPGA-based PCIe board supporting peer-to-peer communication with 
Tesla and Kepler cards~\cite{apenet2010}.
Two nodes of this cluster are used to measure data transfer latency,
one acting as a transmitter and the
other as a receiver.  Data are transferred from the transmitter to the 
receiver, processed on the GPU and sent back to the receiver (see
Fig.~\ref{fig:data_flow}).  
The latency for a complete loop is measured on the transmitter using 
standard \textsc{c} libraries.
\begin{figure}[!t]
\centering
\includegraphics[width=3.5in]{figures/SetUp-general}
\caption{Data flow. Data is sent from the transmitter PC to the
  receiver PC, where it is processed by the GPU before being returned
  to the transmitter PC. The transmitter plays the role of the
  detector as the source of the data and as an upstream trigger
  processor as the data's ultimate sink. The receiver PC plays the
  role of a component in the trigger system. }
\label{fig:data_flow}
\end{figure}
In this setup, the transmitter can represent the detector, as
the source of the data, or an upstream trigger processor, as
the ultimate sink of the data, while the the receiver is the
trigger system: the time to transfer data to the receiver is thus a
rough estimate of the latency to transfer the data from the detector
front-end to the trigger system.

@@

\bibliographystyle{iopart-num}
\bibliography{gpu}

>>>>

\documentclass[journal]{IEEEtran}
\usepackage{graphicx}
\usepackage{subfigure}
% \usepackage[cmex10]{amsmath}
% \usepackage{cite}

\usepackage{hyperref}
% \usepackage{multirow}
% \usepackage[all]{hypcap} % fix links to figs/tabs

@@

% disable the running header on all pages following the title page
\pagestyle{empty}
%
\begin{document}
% paper title
\title{Applications of Many-Core Technologies to On-line Event Reconstruction in High Energy Physics
Experiments}
% Author list
\author{  A.~Gianelle, 
  S.~Amerio, 
  D.~Bastieri, 
  M.~Corvo, 
  W.~Ketchum,
  T.~Liu, 
  A.~Lonardo, 
  D.~Lucchesi,
  S.~Poprocki, 
  R.~Rivera, 
  L.~Tosoratto,
  P.~Vicini
  and 
  P.~Wittich
}
\thanks{Manuscript received November 15, 2013.
This work was supported by the National Science Foundation, the
Department of Energy Office of Science and INFN.}% <-this % stops a space
\thanks{A.~Gianelle, S.~Amerio, D.~Bastieri, M.~Corvo, and D.~Lucchesi are with INFN Padova.}% 
\thanks{W.~Ketchum is with Los Alamos National Laboratory.}%
\thanks{T.~Liu and R.~Rivera are with Fermi National Accelerator Laboratory.}%
\thanks{A.~Lonardo, L.~Tosoratto and P.~Vicini are with INFN Roma}% 
\thanks{S.~Poprocki and P.~Wittich are with Cornell University.}%
% make the title area
\maketitle
% after \maketitle, add the following command to remove the header on the title page
\thispagestyle{empty}
% abstract usually in passive voice
\begin{abstract}
  Interest in parallel architectures applied to real time selections is 
  growing in High Energy Physics (HEP) experiments. In this paper we describe
  performance measurements of Graphic Processing Units (GPUs) and Intel Many
  Integrated Core architecture (MIC) when applied to a typical HEP online task: 
  the selection of events based on the trajectories of charged particles.
  We use as benchmark a scaled-up version of the algorithm used at CDF 
  experiment at Tevatron for online track reconstruction - the SVT algorithm - 
  as a realistic  test-case for low-latency trigger systems using new computing
  architectures for LHC experiment. We examine the
  complexity/performance trade-off in porting existing serial
  algorithms to many-core devices. Measurements of both data processing 
  and data transfer latency are shown, considering different I/O strategies 
  to/from the parallel devices. \end{abstract}

\section{Introduction}
\IEEEPARstart{R}{eal}-time event reconstruction plays a fundamental role in High Energy
Physics (HEP) experiments at hadron colliders.  Reducing the rate of
data to be saved on tape from millions to hundreds of events per
second is critical. To increase the purity of the collected samples,
rate reduction has to be coupled with an initial selection of the most
interesting events.  In a typical hadron collider experiment, the
event rate has to be reduced from tens of MHz to a few kHz.  The
selection system (trigger) is usually organized in multiple levels,
each capable of performing a finer selection on more complex physics
objects describing the event. Trigger systems usually comprise a first
level based on custom hardware, followed by one or two levels usually
based on farms of general purpose processors.  At all levels, latency
is a concern: for a fixed processing time, the faster a decision is
rendered about accepting or rejecting an event improves the purity of
the collected data sample. The possibility of using commercial devices
at a low trigger level is very appealing: they are subject to
continuous performance improvements driven by the consumer market, are
less expensive than dedicated hardware, and are easier to support.
Among the commercial devices, many-core architectures such as Graphic
Processing Units (GPUs)~\cite{bib_gpu} and Intel Many Integrated Core
(MIC)~\cite{bib_intelMIC} are of particular interest for online
selections given their great computing power: the latest
\textsc{nvidia}~\cite{bib_nvidia} GPU architecture, Kepler, exceeds
Teraflop computing power. Moreover, high-level programming
architectures based on \textsc{c/c++} such as
\textsc{cuda}~\cite{bib_cuda} and \textsc{opencl}~\cite{bib_opencl}
make programming these devices more accessible to the general
physicist user.  The goal of this study is to investigate the
strengths and weaknesses of many-core devices when applied in a low
latency environment, with particular emphasis on the data transfer
latency to/from the device and the algorithm latency for processing on
the device in a manner similar to a typical HEP trigger application,
and to understand the cost/complexity ratio of porting legacy serial
code to many-core devices.

@@

\begin{table*}[tbp]
  \caption{Capabilities of the many-core devices used in this study,
    according to the manufacturer's specifications. The first three
    are \textsc{nvidia} GPUs, the MIC 5110P is an Intel Xeon Phi, and the final one is an AMD GPU.}
  \label{tab_hwspecs}
  \centering
  \begin{tabular}{|l|c|c|c|c|c|}
    \hline
    Model & Tesla M2050 & Tesla K20m & GeForce GTX  Titan & MIC 5110P & Radeon HD 7970 \\
    \hline
    \hline
    Performance (SP, GFlops) & 1030 & 3520 & 4500 & 2022 & 3790\\
    Memory bandwidth  (GB/s) & 148 & 208 & 288  & 320 & 264\\   
    Memory size (GB) & 3 & 5 & 6 & 8 & 3\\
    Number of cores & 448 & 2496 & 2688 & 240 & 2048\\
    Clock speed (GHz) & 1.15 & 0.706 & 0.837 & 1.053 & 1.375\\
    \hline
  \end{tabular}
\end{table*}

To measure the data transfer latency we use a computing cluster composed of 
12 identical nodes.  Each node contains a Intel Xeon E6520 2.4 GHz 
CPU and two Tesla M2075 GPU cards. The nodes are connected by InfiniBand 
communication links using Connect-X2 Mellanox or APEnet+ adapters. 
APEnet+ is an FPGA-based PCIe board supporting peer-to-peer communication with 
Tesla and Kepler cards~\cite{apenet2010}.
Two nodes of this cluster are used to measure data transfer latency,
one acting as a transmitter and the
other as a receiver.  Data are transferred from the transmitter to the 
receiver, processed on the GPU and sent back to the receiver (see
Fig.~\ref{fig:data_flow}).  
The latency for a complete loop is measured on the transmitter using 
standard \textsc{c} libraries.
\begin{figure}[tbp]
\centering
\includegraphics[width=3.5in]{figures/SetUp-general}
\caption{Data flow. Data is sent from the transmitter PC to the
  receiver PC, where it is processed by the GPU before being returned
  to the transmitter PC. The transmitter plays the role of the
  detector as the source of the data and as an upstream trigger
  processor as the data's ultimate sink. The receiver PC plays the
  role of a component in the trigger system. }
\label{fig:data_flow}
\end{figure}
In this setup, the transmitter can represent the detector, as
the source of the data, or an upstream trigger processor, as
the ultimate sink of the data, while the the receiver is the
trigger system: the time to transfer data to the receiver is thus a
rough estimate of the latency to transfer the data from the detector
front-end to the trigger system.

@@

\bibliographystyle{IEEEtran}

\bibliography{gpu}
